OptiFit: an improved method for fitting amplicon sequences to existing OTUs

### 2021-12-15 paper/response-to-reviewers.Rmd

---
date: '`r Sys.Date()`'
output:
  pdf_document:
    latex_engine: xelatex
  github_document:
    html_preview: false
header_includes:
  - \usepackage{helvet} % Helvetica font
  - \renewcommand*\familydefault{\sfdefault}
fontsize: 12pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
```

# Response to Reviewers

## Reviewer #1 (Comments for the Author):

**In this manuscript, Sovacool et al. describe a new method for fitting amplicon
sequences to existing datasets. This is an interesting problem that emerges in
microbial ecology: how do you compare new sequence data to existing results
without reprocessing all of the sequences together and creating new de novo
sequences. Early (haha) in microbial ecology, one proposed method of generating
OTUs was to align new data to an existing reference database. This was done by
either a closed (only fitting to the existing data) or open (fitting as much as
possible to the existing data but then generating additional de novo OTUs as
needed) appraoch. Due to many drawbacks, this method is not widely used these
days. However, one benefit it had was the ability to add new data to an existing
set of OTUs. This allowed researchers to make comparisons across projects or
even to conduct meta-type-analyses. In this manuscript, the authors describe a
new method that uses the strengths of modern clustering methods (e.g.,
OptiClust) and the ability to perform cross study comparisons through reference
mapping. The others do a very thorough job of describing the new method,
comparing it to existing methods, exploring the potential user options, and
summarizing everything in a way that gives the reader an understanding of
exactly why they may want to use the new tool, how to use it, and what benefits
it will have.**

**The manuscript is well written and describes a new computational tool without
computer science jargon. I think this manuscript and the method will be very
well received because it does fill a gap in the computation analysis of amplicon
sequences. A have only a few comments and suggestions as outlined below:**

**OTU quality: The main tool for comparing OTU quality in this manuscript is
MCC. This metric is central to de novo OTU clustering in the OptiClust approach,
and I think it greatly benefits amplicon-based approaches. The MCC metric is
briefly described on Lines 54-55. However, since it is so central to the
comparisons in the current study, I wonder if the reader would benefit from a
bit more detail and, I think the reader needs to be reminded about MCC when
interpreting the results. When reviewing Fig 1: the reader should know exactly
why an MCC of 0.91 is better than 0.78 and have a way to conceptualizing the
difference. Also, it may be helpful to remind the reader how quality was
determined (see L172).**

**OptiClust: On lines 78-80 the authors state that "previous studies" found
OptiClust generate the highest quality OTUs, but then they have only a single
reference which is the original paper describing the method (and from the
corresponding authors research group). Full disclosure, I use OptiClust.
However, I think the authors need to either say "A previous study" or provide
additional references (and perhaps one from outside the research group).**

**Computational Resources: Throughout the manuscript, the authors often use
compute time as a measure of the computational resource demand. This is an
important metric, but I also wonder about RAM. Specifically, at the end of the
manuscript the authors state "we recommend using OptiClust for de novo
sequencing over the split strategy with OptiFit since OptiClust is simpler to
execute and performs similarly in terms of both run time and OTU quality"
(L279-281). Earlier they also stated that OptiClust was faster (L244 and L257).
Regardless of time comparisons, I'm also wondering if the OptiFit split strategy
could be an added benefit for individuals that lack computation performance
(e.g., RAM). I don't remember from the OptiClust paper how much RAM is needed or
if it scaled-up with data size, but I thought it was interesting that 10% of a
sequencing run could be used to generate de novo sequences with OptiClust and
then OptiFit could be used to match to other 90% without sacrificing quality.
Could this be a benefit for research groups that may be computationally
limited?**

**Other Reference Based Methods: I think the authors make a good point when the
compare VSEARCH and OptiFit (L261-271). Specifically, I found it interesting to
think about how VSEARCH can allow more dissimilar sequences to be clustered into
the reference OTU. I think this is something that is worth turning into a
stronger statement. For example, since VSEARCH could cluster 3% dissimilar to a
centroid without considering the other sequences being clustered, does that mean
a maximum of 6% dissimilar across all reference clustered sequences is
theoretically possible? In other words, is VSEARCH clustering more in the closed
approach because the within-OTU maximum dissimilarity is actually very high?
This was just a though, but perhaps another way to strengthen the discussion.**

## Reviewer #2 (Comments for the Author):

**The manuscript introduces optiFit, an algorithm that builds on OptiClust that
aims to provide an efficient and robust method to fit amplicon data to existing
OTUs. The authors utilize OptiClust output as their starting list of reference
OTUs, and aim to cluster their new query sequences to the existing output. The
results section is logical and very well written: It describes how the algorithm
works, the benchmarks its performance with various reference databases (SILVA,
GreenGenes, and RDP), compare the results with QIME2 clustering approaches.**

**The reviewer is by no means an expert in computational biology, and merely has
a microbial ecology based background. As such, issues regarding performance
metrics, algorithm strategy, coefficient, etc.. are beyond my expertise. Judging
merely based on the need for such algorithm, the overall strategy, and the
apparent results and benchmarking using reference datasets and databases, I find
the results convincing. The need for such addition to the suite of approaches
offered by MOTHUR is also well explained. I encourage the authors to define what
"high quality OTUs" and "low quality OTUs" means and how they are assessed early
on in the introduction.**

## Reviewer #3 (Comments for the Author):

**In this manuscript, Sovacool et. al. describe OptiFit, a new reference-based
clustering method for assigning (amplicon) read sequences to representative
OTUs. This method (OptiFit) builds upon the concept of OptiClust, a de novo
clustering and OTU picking method previously developed by co-author Westcott.
When carrying out clustering of a given read sequence, OptiClust utilizes
information about all of the sequences within an OTU, instead of only the
representative or centroid sequence for that OTU cluster, an approach used by
many other clustering methods. This enables OptiClust to generate higher quality
OTUs when compared to other de novo methods. OptiFit uses OptiClust's wholistic
method (considering all sequences within a given reference OTU) when performing
reference-based clustering/OTU picking. All query sequences start as their own
OTUs, and the OptiFit algorithm reassigns those sequences to one of the
reference OTUs according to which assignment would maximize the Matthews
Correlation Coefficent (MCC) for the dataset; this approach takes into account
each of the query sequences' similarities to all of the sequences within each
reference OTU. OptiFit is capable of carrying out both open and closed
reference-based clustering.**

**The authors demonstrate OptiFit to be a powerful reference-based clustering
method. The quality of OTU assignments obtained by OptiFit is similar to those
generated by OptiClust, although there is a run-time penalty for OptiFit when
operating open reference mode, although it does not appear to be prohibitive. As
the authors note, the quality of OTU assignments is highly dependent upon the
nature of the samples being studied, as well as the reference databases being
utilized, particularly when carrying out closed reference-based clustering. The
inclusion of an option to perform open reference-based clustering was a
beneficial decision made by the authors. The authors also compared the
performance of OptiFit (and OptiClust) to VSEARCH. OptiFit outperforms VSEARCH
in terms of both speed (execution time) and OTU quality (quantified using median
MCC scores), although VSEARCH is generally able to classify a larger percent of
reads to an OTU in closed reference mode; as the authors note, this is due to
VSEARCH only considering an OTU's centroid sequence, as opposed to all of the
sequences within that particular OTU, making OptiFit more selective.**

**The OptiFit algorithm is a natural extension of the OptiClust approach to
sequence classification, albeit for situations where a reference-based approach
is either required or beneficial. One of the more appealing aspects of OptiFit
is that it does not change OTU assignments for samples/data that have been
already analyzed when adding new information to a dataset. This functionality
would be particularly useful for investigators who are carrying out
meta-analysis type studies or who have very large datasets to which new samples
are added regularly and reanalysis would be computationally burdensome.**

**A general question I have is whether the authors have considered suggesting
that users make use of the data-splitting approach that they outline in their
own paper. OTU classification is a duel task problem (identifying OTUs in a
dataset and assigning each read sequence to an appropriate OTU). With de novo
classification approaches the same data is used to both identify and classify
the reads. From my perspective, the development of OptiFit makes a perfect
partner for OptiClust in a Train/Test (Train/Classify) splitting situation,
where the users would split data, train using OptiClust and then classify the
remaining reads using OptiFit. Of course, pursuing this would be well beyond the
scope of this manuscript, but the authors may consider discussing such a
possibility for future work.**

**The manuscript is generally clear, concise, and well written. There are only
so many times that one can write "clustered"/"clustering" (or other variants) in
a sentence or paragraph and still have it make sense. I can tell (and
appreciate) that the authors took great care in the writing process to properly
convey their message.**

**Specific comments:**

**Line 4 (Abstract): The sentence discussing OptiClust appears to be tangential
to the opening of the abstract. Perhaps it would be better to add this context
further down in the abstract (line 16-ish).**

**Line 83-85: "While other tools represent reference OTUs with a single
sequence, OptiFit 84 uses multiple sequences in existing OTUs as the reference
and fits new sequences to 85 those reference OTUs." Should "multiple" be "all"
in this sentence?**

**Lines 129 - 131: Please include references for Greengenes and SILVA, and
indicate what versions of all three (Greengenes, SILVA, and RDP) were used, even
though this information is included in the Methods section.**

**Lines 136-138: "Clustering sequences to ..." I think it would serve the reader
to reiterate/emphasize that OptiFit was used for clustering, i.e. "Clustering
query sequences with OptiFit to ...".**

**Lines 136-138: I suggest adding the corresponding Figure 3 labels (in
parentheses), similar to what was done for the "self-split" tests described in
Lines 196-198. This will help guide the reader, as there is quite a bit of
information being displayed in Figure 3 as well as what is being described in
this paragraph.**

**Lines 141-142: "Thus, open reference OptiFit produced OTUs of very similar
quality as de novo clustering, ..." please add "with OptiClust" to be more
specific about what the comparison being made is.**

**Figure 1: W-Z are designated as the query sequences in the caption. Z does not
appear in the distance threshold table at the top of the figure and is not
assigned to an OTU at the end of Cycle 4. From Line 115 ("Alternatively, a
sequence will remained unassigned...") the reader can infer that sequence Z is
different enough from the others that it was designated a singleton (outside of
the distance threshold discussed in lines 111-113). If I'm correct, I encourage
the authors to make this point explicitly in Figure 1's caption.**

**Figure 3: I suggest changing the label for "de novo" to "OptiClust (de novo)"
or similar to emphasize that OptiClust was being used for those benchmark
tests.**

**Figure 3: The second sentence in the caption ("Each dataset...") is difficult
to follow due to its structure. I suggest rewriting to make it clear that de
novo clustering was performed with OptiClust, and 4 different reference based
trials were performed (self-split (referencing the OptiClust-ed reference
subset), referencing GG, SILVA, & RDP).**

### 2021-12-15 paper/response-to-reviewers.md

2021-12-15

# Response to Reviewers

## Reviewer #1 (Comments for the Author):

**In this manuscript, Sovacool et al. describe a new method for fitting
amplicon sequences to existing datasets. This is an interesting problem
that emerges in microbial ecology: how do you compare new sequence data
to existing results without reprocessing all of the sequences together
and creating new de novo sequences. Early (haha) in microbial ecology,
one proposed method of generating OTUs was to align new data to an
existing reference database. This was done by either a closed (only
fitting to the existing data) or open (fitting as much as possible to
the existing data but then generating additional de novo OTUs as needed)
appraoch. Due to many drawbacks, this method is not widely used these
days. However, one benefit it had was the ability to add new data to an
existing set of OTUs. This allowed researchers to make comparisons
across projects or even to conduct meta-type-analyses. In this
manuscript, the authors describe a new method that uses the strengths of
modern clustering methods (e.g., OptiClust) and the ability to perform
cross study comparisons through reference mapping. The others do a very
thorough job of describing the new method, comparing it to existing
methods, exploring the potential user options, and summarizing
everything in a way that gives the reader an understanding of exactly
why they may want to use the new tool, how to use it, and what benefits
it will have.**

**The manuscript is well written and describes a new computational tool
without computer science jargon. I think this manuscript and the method
will be very well received because it does fill a gap in the computation
analysis of amplicon sequences. A have only a few comments and
suggestions as outlined below:**

**OTU quality: The main tool for comparing OTU quality in this
manuscript is MCC. This metric is central to de novo OTU clustering in
the OptiClust approach, and I think it greatly benefits amplicon-based
approaches. The MCC metric is briefly described on Lines 54-55. However,
since it is so central to the comparisons in the current study, I wonder
if the reader would benefit from a bit more detail and, I think the
reader needs to be reminded about MCC when interpreting the results.
When reviewing Fig 1: the reader should know exactly why an MCC of 0.91
is better than 0.78 and have a way to conceptualizing the difference.
Also, it may be helpful to remind the reader how quality was determined
(see L172).**

**OptiClust: On lines 78-80 the authors state that “previous studies”
found OptiClust generate the highest quality OTUs, but then they have
only a single reference which is the original paper describing the
method (and from the corresponding authors research group). Full
disclosure, I use OptiClust. However, I think the authors need to either
say “A previous study” or provide additional references (and perhaps one
from outside the research group).**

**Computational Resources: Throughout the manuscript, the authors often
use compute time as a measure of the computational resource demand. This
is an important metric, but I also wonder about RAM. Specifically, at
the end of the manuscript the authors state “we recommend using
OptiClust for de novo sequencing over the split strategy with OptiFit
since OptiClust is simpler to execute and performs similarly in terms of
both run time and OTU quality” (L279-281). Earlier they also stated that
OptiClust was faster (L244 and L257). Regardless of time comparisons,
I’m also wondering if the OptiFit split strategy could be an added
benefit for individuals that lack computation performance (e.g., RAM). I
don’t remember from the OptiClust paper how much RAM is needed or if it
scaled-up with data size, but I thought it was interesting that 10% of a
sequencing run could be used to generate de novo sequences with
OptiClust and then OptiFit could be used to match to other 90% without
sacrificing quality. Could this be a benefit for research groups that
may be computationally limited?**

**Other Reference Based Methods: I think the authors make a good point
when the compare VSEARCH and OptiFit (L261-271). Specifically, I found
it interesting to think about how VSEARCH can allow more dissimilar
sequences to be clustered into the reference OTU. I think this is
something that is worth turning into a stronger statement. For example,
since VSEARCH could cluster 3% dissimilar to a centroid without
considering the other sequences being clustered, does that mean a
maximum of 6% dissimilar across all reference clustered sequences is
theoretically possible? In other words, is VSEARCH clustering more in
the closed approach because the within-OTU maximum dissimilarity is
actually very high? This was just a though, but perhaps another way to
strengthen the discussion.**

## Reviewer #2 (Comments for the Author):

**The manuscript introduces optiFit, an algorithm that builds on
OptiClust that aims to provide an efficient and robust method to fit
amplicon data to existing OTUs. The authors utilize OptiClust output as
their starting list of reference OTUs, and aim to cluster their new
query sequences to the existing output. The results section is logical
and very well written: It describes how the algorithm works, the
benchmarks its performance with various reference databases (SILVA,
GreenGenes, and RDP), compare the results with QIME2 clustering
approaches.**

**The reviewer is by no means an expert in computational biology, and
merely has a microbial ecology based background. As such, issues
regarding performance metrics, algorithm strategy, coefficient, etc..
are beyond my expertise. Judging merely based on the need for such
algorithm, the overall strategy, and the apparent results and
benchmarking using reference datasets and databases, I find the results
convincing. The need for such addition to the suite of approaches
offered by MOTHUR is also well explained. I encourage the authors to
define what “high quality OTUs” and “low quality OTUs” means and how
they are assessed early on in the introduction.**

## Reviewer #3 (Comments for the Author):

**In this manuscript, Sovacool et. al. describe OptiFit, a new
reference-based clustering method for assigning (amplicon) read
sequences to representative OTUs. This method (OptiFit) builds upon the
concept of OptiClust, a de novo clustering and OTU picking method
previously developed by co-author Westcott. When carrying out clustering
of a given read sequence, OptiClust utilizes information about all of
the sequences within an OTU, instead of only the representative or
centroid sequence for that OTU cluster, an approach used by many other
clustering methods. This enables OptiClust to generate higher quality
OTUs when compared to other de novo methods. OptiFit uses OptiClust’s
wholistic method (considering all sequences within a given reference
OTU) when performing reference-based clustering/OTU picking. All query
sequences start as their own OTUs, and the OptiFit algorithm reassigns
those sequences to one of the reference OTUs according to which
assignment would maximize the Matthews Correlation Coefficent (MCC) for
the dataset; this approach takes into account each of the query
sequences’ similarities to all of the sequences within each reference
OTU. OptiFit is capable of carrying out both open and closed
reference-based clustering.**

**The authors demonstrate OptiFit to be a powerful reference-based
clustering method. The quality of OTU assignments obtained by OptiFit is
similar to those generated by OptiClust, although there is a run-time
penalty for OptiFit when operating open reference mode, although it does
not appear to be prohibitive. As the authors note, the quality of OTU
assignments is highly dependent upon the nature of the samples being
studied, as well as the reference databases being utilized, particularly
when carrying out closed reference-based clustering. The inclusion of an
option to perform open reference-based clustering was a beneficial
decision made by the authors. The authors also compared the performance
of OptiFit (and OptiClust) to VSEARCH. OptiFit outperforms VSEARCH in
terms of both speed (execution time) and OTU quality (quantified using
median MCC scores), although VSEARCH is generally able to classify a
larger percent of reads to an OTU in closed reference mode; as the
authors note, this is due to VSEARCH only considering an OTU’s centroid
sequence, as opposed to all of the sequences within that particular OTU,
making OptiFit more selective.**

**The OptiFit algorithm is a natural extension of the OptiClust approach
to sequence classification, albeit for situations where a
reference-based approach is either required or beneficial. One of the
more appealing aspects of OptiFit is that it does not change OTU
assignments for samples/data that have been already analyzed when adding
new information to a dataset. This functionality would be particularly
useful for investigators who are carrying out meta-analysis type studies
or who have very large datasets to which new samples are added regularly
and reanalysis would be computationally burdensome.**

**A general question I have is whether the authors have considered
suggesting that users make use of the data-splitting approach that they
outline in their own paper. OTU classification is a duel task problem
(identifying OTUs in a dataset and assigning each read sequence to an
appropriate OTU). With de novo classification approaches the same data
is used to both identify and classify the reads. From my perspective,
the development of OptiFit makes a perfect partner for OptiClust in a
Train/Test (Train/Classify) splitting situation, where the users would
split data, train using OptiClust and then classify the remaining reads
using OptiFit. Of course, pursuing this would be well beyond the scope
of this manuscript, but the authors may consider discussing such a
possibility for future work.**

**The manuscript is generally clear, concise, and well written. There
are only so many times that one can write “clustered”/“clustering” (or
other variants) in a sentence or paragraph and still have it make sense.
I can tell (and appreciate) that the authors took great care in the
writing process to properly convey their message.**

**Specific comments:**

**Line 4 (Abstract): The sentence discussing OptiClust appears to be
tangential to the opening of the abstract. Perhaps it would be better to
add this context further down in the abstract (line 16-ish).**

**Line 83-85: “While other tools represent reference OTUs with a single
sequence, OptiFit 84 uses multiple sequences in existing OTUs as the
reference and fits new sequences to 85 those reference OTUs.” Should
“multiple” be “all” in this sentence?**

**Lines 129 - 131: Please include references for Greengenes and SILVA,
and indicate what versions of all three (Greengenes, SILVA, and RDP)
were used, even though this information is included in the Methods
section.**

**Lines 136-138: “Clustering sequences to …” I think it would serve the
reader to reiterate/emphasize that OptiFit was used for clustering,
i.e. “Clustering query sequences with OptiFit to …”.**

**Lines 136-138: I suggest adding the corresponding Figure 3 labels (in
parentheses), similar to what was done for the “self-split” tests
described in Lines 196-198. This will help guide the reader, as there is
quite a bit of information being displayed in Figure 3 as well as what
is being described in this paragraph.**

**Lines 141-142: “Thus, open reference OptiFit produced OTUs of very
similar quality as de novo clustering, …” please add “with OptiClust” to
be more specific about what the comparison being made is.**

**Figure 1: W-Z are designated as the query sequences in the caption. Z
does not appear in the distance threshold table at the top of the figure
and is not assigned to an OTU at the end of Cycle 4. From Line 115
(“Alternatively, a sequence will remained unassigned…”) the reader can
infer that sequence Z is different enough from the others that it was
designated a singleton (outside of the distance threshold discussed in
lines 111-113). If I’m correct, I encourage the authors to make this
point explicitly in Figure 1’s caption.**

**Figure 3: I suggest changing the label for “de novo” to “OptiClust (de
novo)” or similar to emphasize that OptiClust was being used for those
benchmark tests.**

**Figure 3: The second sentence in the caption (“Each dataset…”) is
difficult to follow due to its structure. I suggest rewriting to make it
clear that de novo clustering was performed with OptiClust, and 4
different reference based trials were performed (self-split (referencing
the OptiClust-ed reference subset), referencing GG, SILVA, & RDP).**

### 2021-12-15 paper/paper.Rmd
Bold reviewer comments & use R Markdown

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_knit$set(base.dir = here::here(), 
                     base.url = "../")
knitr::opts_chunk$set(fig.path = 'figures/',
                      dpi = 300)
```

>>>>

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(fig.path = 'figures/',
                      dpi = 300)
```
